### 2011-12-01 papers/smectics/smectics.tex
Update OH
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1492 40876441-3496-0410-a613-a2fca35b2ae2

\documentclass[8.5pt,twoside,twocolumn]{article}
\oddsidemargin -1.2cm
\evensidemargin -1.2cm
\textwidth 18cm
\headheight 1.0in
\topmargin -3.5cm
\textheight 22cm
\usepackage[super,sort&compress,comma]{natbib} 
\usepackage{mhchem}
\usepackage{times,mathptm}
\usepackage{sectsty}
\usepackage{balance} 
\usepackage{graphicx} %eps figures can be used instead
\usepackage{lastpage}
\usepackage[format=plain,justification=raggedright,singlelinecheck=false,font=small,labelfont=bf,labelsep=space]{caption} 
\usepackage{fancyhdr}
\usepackage{bm}
\usepackage{float}
%\usepackage{rsc}

@@

\footnotetext{\textit{$^{a}$ Centre for Computational Science, University College London, INSERT ADDRESS HERE, UK}}
\footnotetext{\textit{$^{b}$ Edinburgh Parallel Computing Centre, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}}
\footnotetext{\textit{$^{c}$ SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}}

@@

SADLY, I THINK IT WOULD BE BEST TO REWRITE THIS PAPER WITH THE FLOW DIRECTION ALONG X, THE GRADIENT ALONG Y AND THE VORTICITY ALONG Z AS IS COMMONLY USED IN THE RHEOLOGICAL LITERATURE! I HAVE NOT STARTED TO MAKE THIS CHANGE AS IT WOULD CONFUSE THINGS TOO MUCH AND THE JOB REQUIRES CARE. ALSO THE CHEMICAL PRESSURE AND THE STRESS HAVE OPPOSITE SIGNS YET THE FIGURES SHOW THEM HAVING THE SAME SIGN. WHAT HAS HAPPENED?

@@

The flow behaviour (rheology) of smectic liquid crystals affects many industrial and manufacturing processes; conversely, rheological measurements provide important probes of their macroscopic material properties.
For a simple fluid the response to steady applied stress is quantified in the linear (Newtonian) regime by a single viscosity. Lamellar phases are anisotropic, and thus in principle have several viscosities as well as an elastic modulus related to layer compression. However, these linear response parameters become somewhat irrelevant when defects are present: unless their density is very low, the defects soon dominate the stress. Moreover, since the defect pattern is strongly dependent on flow rate, the rheology is generically nonlinear. On shearing, the effective viscosity (ratio of shear stress $\sigma_{xy}$  to strain rate $\dot\gamma$) may either increase or decrease, corresponding
to shear thickening or shear thinning behaviour [Larson book]. Also if a defect texture is present in the system at rest, this may show a viscosity divergence at small $\dot \gamma$ --- i.e., a yield stress.
Finally, in common with other complex fluids, lamellar phases can undergo transitions into heterogeneous states, for instance to form shear bands (layers of unequal $\sigma_{xy}$ or $\dot\gamma$) [P D Olmsted, Rheologica Acta 47, 283-300 (2008)].

Lamellar liquid crystals provide an important paradigm for non-Newtonian fluids in which the microstructure of the material is strongly coupled to the flow. Crucially, the defect texture may depend on past history -- for instance, the strain rate in any prior flow that was suddenly switched off. This rheological hysteresis means that the precise flow protocol, including any pre-shearing applied prior to starting the experiment, needs to be specified. Such hysteresis can arise in polydomain liquid crystals of various types (not just smectics) such as nematics in which the existence and evolution of regions with misaligned directors creates rheological nonlinearity. A closely related example is the so-called `region I flow' in polymeric nematic liquid crystals, where it is thought that a defect network arises, with enhanced viscosity, but with
marked shear thinning as this structure is sheared out at high flow rates [Larson book -- or Davide to supply references].

An intriguing complication in addressing the dynamics of layered systems
is permeation,
where individual molecules in the fluid translate while keeping the
layerwise order intact. In steady state, permeation leads to increased dissipation
and hence to a large viscosity in any geometry where flow necessitates such motion [deGennes book]. (A well-known example of permeation flow
is found in cholesterics which are sheared or pushed along the direction
of the cholesteric helix.) Permeation is often involved in defect motion such as the growth or shrinkage of incomplete layers at the sites of edge dislocations, and may also be implicated in other forms of topological reconstruction. Such reconstruction arises under shear, sometimes to an extreme degree, for instance in the shear-induced transition from an ordered lamellar stack to an onion phase in lyotropic smectics \cite{Diat93,Panizza96,Iwashita07}[ADD; M Gradzielski, JPCM 15, R655-R697 (2003)]. (The onion phase comprises a space-filling packing of multilayer vesicles; under some conditions similar defect textures can arise without shear \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04}.) The mechanism underlying such global reconstructions under shear of the layer organization remains unclear. Most theoretical predictions have focussed on buckling and other instabilities of perfectly ordered smectics as the initial trigger for onion formation [A G Zilman and R Granek, EPJB 11, 593-608 (1999), L Courbin, J P Delville, J Rouch and P Panizza, PRL 89, 148305 (2002)]\cite{Guo2002,Soddemann2004, Stewart2009} but the pathway may also strongly involve the physics of defect motion [J Zipfel, J Berghause, P Lindner, W Richtering, J Phys Chem B 103, 2841-2849 (1999), A Leon, D Bonn, J Meunier, A Al-Kahlwaji, O Greffer, H Kellay, PRL 84, 1355-1338 (2000)]. Although we do not see onions in the simulations reported here, in two dimensions we do see a transition from a well ordered phase to a defect-rich phase, whose possible link to onion formation deserves to be explored in future work. Another interesting avenue connects with the observation, in block copolymers and some lyotropic surfactant systems, of a shear-induced
phase transitions from a homogeneous state into to ordered phases
like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}. However, theory suggests that thermal noise (causing local fluctuations about the mean state of order) play a central role in this transition~\cite{Cates89}. In the present work we focus on the simplest case in which such thermal fluctuations are ignored, and the time evolution of the microstructure is deterministic (though certainly chaotic and unpredictable
once defects are present).

@@

In this work, we try to overcome these technical obstacles, presenting systematic results for large two dimensional simulations with fully periodic boundary conditions.
Guided by those results, we additionally report some very large three dimensional simulations at selected points in the parameter space. These 3D simulations are, we believe, of unprecedented scale for lamellar phases, although even larger ones have recently been used to address the rheology of cubic liquid crystals~\cite{Saksena09}. [ADD: R S Saksena and P V Coveney, Proc. Roy. Soc. A. 465, 1977-2002 (2009)).]

We describe smectics below via a widely-used free energy functional [G Gompper and M Schick, Phase Transitions and Critical Phenomena vol 16, C Domb and J L Lebowitz, Eds., Academic Press NY 1994; then cite all the lamellar simulation papers using this or similar functionals]. This provides a robust generic framework for addressing lamellar ordering without being tied to any specific class of materials (although we do become more specific when estimating suitable parameters for the simulations). The free energy is formulated in terms of a scalar order parameter $\phi({\bf r})$. This can be viewed as describing a composition variable in a binary fluid system stabilized by added amphiphiles; one can then speak of alternating layers of fluid and interfaces between these layers. A similar interpretation applies to block copolymers where the composition variable describes the relative local density of the two blocks. However, the same order parameter could equally describe the local number density of the midpoints of a rodlike mesogenic molecule (measured relative to the mean) which can develop a periodic density wave that represents the smectic state.

The governing equations in our model are a convection-diffusion equation
for the scalar order parameter, and the Navier-Stokes equation describing
the hydrodynamics of fluid flow. The stress tensor for the fluid contains not only a Newtonian viscous term but extra contributions arising from local deviations of the order parameter from its state of least free energy. Diffusive relaxation of the order parameter then causes additional dissipation, resulting in an effective viscosity that can be much higher than the intrinsic Newtonian contribution.
Our dynamical equations are solved using a combination of a finite
difference scheme for the convection-diffusion equation, and the
lattice Boltzmann method (LBM)~\cite{Succi} for the Navier-Stokes equation
[CITE OUR PREVIOUS PAPERS USING HYBRID LB, ALSO THAT OF GONELLA ET AL WHO FIRST USED IT FOR SMECTICS].
We use sliding periodic (Lees-Edwards) boundary conditions for
both the convection-diffusion and Navier-Stokes equations. Note that while these boundary conditions impose a fixed integrated strain rate across the whole system they do not impose a uniform strain rate locally; shear bands can therefore arise if the system so chooses. (As it happens, we do not observe shear bands in the range of shear rates we address here.)

@@

% KS: this following paragraph doesn't quite make sense to me yet.

Our main finding is that, in two dimensions at least, sheared smectics can show two completely different types of rheological response. At modest rates
of shear, where thermodynamic forces are relatively strong compared with
flow effects,  shear quickly aligns the system into the perfect lamellar phase preferred by thermodynamics,
with only a few defects. The rheology in this regime is quasi-Newtonian, with shear stress linear in the applied shear rate. (There is however a large first normal stress difference.) At higher shear rates, larger thermodynamic forces are generated as layers fail to keep up with the imposed flow. This leads to the nucleation of many new defects, balanced in steady state by annihilation when defects of opposite sign encounter each other. The strain rate $\dot\gamma_c$ needed to enter this nonlinear flow regime depends on thermodynamic and kinetic parameters; a comprehensive exploration of parameter space is not possible, but we give a scaling argument which relates $\dot\gamma_c$ to those parameters through an effective capillary number, formed as the ratio of viscous to thermodynamic stress.
Within the defect proliferation regime, the macroscopic rheology is generically nonlinear, and shows shear thinning. The shear stress appears to vary as a sublinear power law, $\sigma_{xy}\propto\dot{\gamma}^m$, with $m\simeq 0.87$ for typical parameter values, although the dependence of $m$ on the parameters precludes interpretation of this as a universal scaling law.
Such a behaviour is qualitatively reminiscent of the one found experimentally in [REFERENCE?], where the numerical value of $m$ was however different. In this strongly non-Newtonian regime we also find that the rheology is history dependent.

@@

The equilibrium phase in a system of volume $V$ is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimises the Landau-Brazovskii free energy functional \cite{Brazovskii75} [ADD: Gommper and Schick].
%
\begin{eqnarray}
{\cal F}[\phi] &=& \int d^dx f(\phi,\nabla\phi,\nabla^2\phi) = V\bar f\\ &=& \int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}\nonumber\label{FE-functional}
\end{eqnarray}
%
In minimizing this free energy at equilibrium, we ignore thermal fluctuations (treated in \cite{Brazovskii75,Cates89}), thereby adopting a mean-field approximation as in previous literature [Kendon et al JFM, relevant Gonnella papers]. This should be a good approximation far from the ordering transition at which $\phi$ first becomes nonuniform. The uniform state has $\phi = 0$ and hence ${\cal F}= 0$.

The parameter range relevant to lamellar phases has $\kappa < 0$ and $c>0$. Then, if we suppress the quartic term, the free energy density of a sinusoidal lamellar state $\phi({\bf r}) = \phi_k\cos({\bf k}.{\bf r})$ obeys $\bar f = (a+\kappa k^2+ c k^4)\phi_k^2/4$. This is minimized by choosing $k=q\equiv\sqrt{|\kappa|/2c}$; the free energy density is then $\bar f = (a-\kappa^2/4c)\phi_q^2/4$ so that ordering will occur for $a<a_c=\kappa^2/4c$. The amplitude $\phi_q$ is then unbounded without the (positive) quartic term, since $\bar f$ can in that manner become infinitely negative. Restoring the quartic term now causes saturation at $\phi_q\sim \sqrt{(a_c-a)/b}$, creating also nonlinear harmonics in the density profile. The final free energy density obeys $\bar f \simeq (a_c-a)^2/b$. By similar arguments, the characteristic elastic modulus associated with microstructural deformations is
%
\begin{equation}
\bar g\sim \frac{(a_c-a)a_c}{4b}
\label{barg}
\end{equation}
%
which is the energy density of the gradient terms only. The terms involving $\phi^2$ and $\phi^4$ in Eq.~\ref{FE-functional} are unaltered by affine deformations which conserve density locally; hence those terms do not contribute to $\bar g$.

The dynamics of the order parameter is governed by an advection-diffusion equation with a diffusive current of $\phi$ dependent on the chemical potential $\mu$, the functional derivative of ${\cal F}$ with respect to $\phi$, as follows:
%
\begin{equation}\label{OPEOM}
\partial_t \phi +{\nabla}({\bm u}\phi) = M \nabla^2 \frac{\delta \cal F}{\delta \phi}=M \nabla^2 \mu 
\end{equation}
%
Here $M$ is a mobility coefficient. A wavevector-dependent linearized collective diffusivity $D_k$ can then be defined via $\dot\phi_k = -k^2D_k\phi_k$, whose value at the ordering wavevector obeys $D_q = M(a-a_c)/2$ [CONFIRM] (this is negative, signifying the instability). In contrast, an estimate of the molecular diffusivity can be made by considering zero wavevector, setting $b = -a$ (see below) and examining the relaxation equation for small deviations around the uniformly ordered state at $\phi = \pm 1$. This gives $\dot\phi = 2bM\nabla^2\phi$ so that the molecular diffusivity is of order $\tilde D = 2Mb$.

Treating the fluid as incompressible, the evolution of the velocity ${\bm u}$ obeys the Navier-Stokes equation
%
\begin{equation}\label{NSE}
\rho(\partial_t u_\alpha + (u_\beta \nabla_\beta) u_\alpha) = \partial_\beta \sigma_{\alpha \beta}. 
\end{equation}
%
where $\rho$ is the mass density. The stress tensor in this equation reads
%
\begin{equation}
\sigma_{\alpha \beta}=\eta (\partial_\alpha u_\beta + \partial_\beta u_\alpha)-p\delta_{\alpha\beta} - P_{\alpha \beta}. \label{fullstress}
\end{equation}
%
This consists of a Newtonian hydrodynamic part, which is proportional to the background viscosity $\eta$ and the symmetrized velocity gradient, a hydrostatic pressure term, which includes an isotropic stress contribution from the order parameter 
%
\begin{eqnarray}
\Delta p&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\nabla}\phi)^2\right)\nonumber\\
& &\quad+c\left(\phi(\nabla^2)^2\phi+\frac{1}{2}(\nabla^2\phi)^2+\partial_{\gamma}\phi\partial_\gamma(\nabla^2\phi)\right)
\end{eqnarray}
%
and finally another term consisting of the order-parameter contribution to the deviatoric stress
\cite{Evans79} [REFERENCE]
%
\begin{equation}
P_{\alpha \beta}=\kappa\partial_\alpha\phi\partial_\beta\phi -c \left(\partial_\alpha\phi\partial_\beta(\nabla^2\phi)+\partial_\beta\phi\partial_\alpha(\nabla^2\phi)\right).
\end{equation}
CONFIRM SIGN.

@@

To calibrate the LB length unit we equate $L$ the lamellar wavelength with that of a typical thermotropic smectic: $10$ LBU $= L = 4$nm. Hence the LB length unit is about $0.4$nm; and 1m$= 2.5\times 10^{10}$LBU. Free energy density parameters such as $a_c \simeq 10^{-4}$LBU should then be of order $\alpha^2k_BT/L^3$ where $\alpha =10$ is a molecular aspect ratio. Thus the LB unit of energy density is about $6\times 10^{10}$Pa, i.e. 1 Pa$= 1.6 \times 10^{-11}$ LBU.

@@

As detailed previously, an estimate of the molecular diffusivity is $\tilde D = 2Mb$; we set $M = 0.25$LBU throughout this study, so that $\tilde D$ lies in the range $2.5\times 10^{-5} - 2.5\times 10^{-4}$LBU. Taking a mid-range value of  $\tilde D = 10^{-4}$LBU and using the above conversion factors (1 m$= 2.5\times 10^9$LBU, 1s $= 6\times 10^{11}$LBU) gives $D\simeq 10^{-11}$m$^2$s$^{-1}$ which is a reasonable mesogenic value.

Recall that the mass density of our fluid is unity in LBU. However, combining the above conversion factors for length,  energy density, and time shows the physical and simulation units of mass density to be related via 1 kg m$^{-3} \equiv 1$ Pa s$^2$ m$^{-2} \simeq 10^{-8}$LBU. Thus our simulations address a fluid of mass density $10^8$kg m$^{-3}$ which is $10^5$ times larger than the true value for a typical material. The reason for this is that the LB algorithm uses fluid inertia to update the dynamics and the discretization should therefore involve replacing the real fluid with the densest one possible, while ensuring that the Reynolds number $Re \sim \rho\dot\gamma L^2/\eta$ (which for real smectics is exceedingly small) remains small enough for the simulation to faithfully describe the low $Re$ limit. As discussed elsewhere [CODEF,SOFT MATTER] $Re$ values up to about 0.1 are then admissable; our simulations comply with this requirement.

Finally, our shear rates lie in the range 
$10^{-5}-10^{-4}$LBU, so that the (dimensionless) Peclet number, $Pe =\dot\gamma L^2/\tilde D$, which expresses the relative importance of advective to diffusive transport, lies in the range $20-1600$ CONFIRM: WRONG VALUES GIVEN PREVIOUSLY?. These are moderate to high values where significant structural rearrangement can be expected. However, for thermotropic smectics, given our time unit of 1LBU$=1.6\times 10^{-12}$s, they correspond to strain rates of order $10^6$s$^{-1}$. This lies well above the range typically addressable with rheological experiments, but far below that addressed recently in molecular dynamics simulations [Guo,Sodderman,Stewart].

>>>>

\documentclass[8.5pt,twoside,twocolumn]{article}
\oddsidemargin -1.2cm
\evensidemargin -1.2cm
\textwidth 18cm
\headheight 1.0in
\topmargin -3.5cm
\textheight 22cm
\usepackage[super,sort&compress,comma]{natbib} 
\usepackage{mhchem}
\usepackage{times,mathptm}
\usepackage{sectsty}
\usepackage{balance} 
\usepackage{graphicx} %eps figures can be used instead
\usepackage{lastpage}
\usepackage[format=plain,justification=raggedright,singlelinecheck=false,font=small,labelfont=bf,labelsep=space]{caption} 
\usepackage{fancyhdr}
\usepackage{bm}
\usepackage{float}

@@

\footnotetext{\textit{$^{a}$ Centre for Computational Science, University College London, 20 Gordon Street, London WC1H 0AJ, UK}}
\footnotetext{\textit{$^{b}$ Edinburgh Parallel Computing Centre, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}}
\footnotetext{\textit{$^{c}$ SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}}

@@

@@

The flow behaviour (rheology) of smectic liquid crystals affects many industrial and manufacturing processes; conversely, rheological measurements provide important probes of their macroscopic material properties.
For a simple fluid the response to steady applied stress is quantified in the linear (Newtonian) regime by a single viscosity. Lamellar phases are anisotropic, and thus in principle have several viscosities as well as an elastic modulus related to layer compression. However, these linear response parameters become somewhat irrelevant when defects are present: unless their density is very low, the defects soon dominate the stress. Moreover, since the defect pattern is strongly dependent on flow rate, the rheology is generically nonlinear. On shearing, the effective viscosity (ratio of shear stress $\sigma_{xy}$  to strain rate $\dot\gamma$) may either increase or decrease, corresponding
to shear thickening or shear thinning behaviour [Larson book]. Also if a defect texture is present in the system at rest, this may show a viscosity divergence at small $\dot \gamma$ --- i.e., a yield stress.
Finally, in common with other complex fluids, lamellar phases can undergo transitions into heterogeneous states, for instance to form shear bands (layers of unequal $\sigma_{xy}$ or $\dot\gamma$) \cite{Olmsted08}.

Lamellar liquid crystals provide an important paradigm for non-Newtonian fluids in which the microstructure of the material is strongly coupled to the flow. Crucially, the defect texture may depend on past history -- for instance, the strain rate in any prior flow that was suddenly switched off. This rheological hysteresis means that the precise flow protocol, including any pre-shearing applied prior to starting the experiment, needs to be specified. Such hysteresis can arise in polydomain liquid crystals of various types (not just smectics) such as nematics in which the existence and evolution of regions with misaligned directors creates rheological nonlinearity. A closely related example is the so-called `region I flow' in polymeric nematic liquid crystals, where it is thought that a defect network arises, with enhanced viscosity, but with
marked shear thinning as this structure is sheared out at high flow rates \cite{Larson}.

An intriguing complication in addressing the dynamics of layered systems
is permeation,
where individual molecules in the fluid translate while keeping the
layerwise order intact. In steady state, permeation leads to increased dissipation
and hence to a large viscosity in any geometry where flow necessitates such motion \cite{deGennes}. (A well-known example of permeation flow
is found in cholesterics which are sheared or pushed along the direction
of the cholesteric helix.) Permeation is often involved in defect motion such as the growth or shrinkage of incomplete layers at the sites of edge dislocations, and may also be implicated in other forms of topological reconstruction. Such reconstruction arises under shear, sometimes to an extreme degree, for instance in the shear-induced transition from an ordered lamellar stack to an onion phase in lyotropic smectics \cite{Diat93,Panizza96,Iwashita07,Gradzielski03}. (The onion phase comprises a space-filling packing of multilayer vesicles; under some conditions similar defect textures can arise without shear \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04}.) The mechanism underlying such global reconstructions under shear of the layer organization remains unclear. Most theoretical predictions have focussed on buckling and other instabilities of perfectly ordered smectics as the initial trigger for onion formation \cite{Zilman99,Courbin02,Guo2002,Soddemann2004, Stewart2009} but the pathway may also strongly involve the physics of defect motion \cite{Zipfel99, Leon00}. Although we do not see onions in the simulations reported here, in two dimensions we do see a transition from a well ordered phase to a defect-rich phase, whose possible link to onion formation deserves to be explored in future work. Another interesting avenue connects with the observation, in block copolymers and some lyotropic surfactant systems, of a shear-induced
phase transitions from a homogeneous state into to ordered phases
like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}. However, theory suggests that thermal noise (causing local fluctuations about the mean state of order) play a central role in this transition~\cite{Cates89}. In the present work we focus on the simplest case in which such thermal fluctuations are ignored, and the time evolution of the microstructure is deterministic (though certainly chaotic and unpredictable
once defects are present).

@@

In this work, we try to overcome these technical obstacles, presenting systematic results for large two dimensional simulations with fully periodic boundary conditions.
Guided by those results, we additionally report some very large three dimensional simulations at selected points in the parameter space. These 3D simulations are, we believe, of unprecedented scale for lamellar phases, although even larger ones have recently been used to address the rheology of cubic liquid crystals \cite{Saksena09a, Saksena09b}.

We describe smectics below via a widely-used free energy functional \cite{Gompper, Gonnella97, Xu06b, Cencini07}. This provides a robust generic framework for addressing lamellar ordering without being tied to any specific class of materials (although we do become more specific when estimating suitable parameters for the simulations). The free energy is formulated in terms of a scalar order parameter $\phi({\bf r})$. This can be viewed as describing a composition variable in a binary fluid system stabilized by added amphiphiles; one can then speak of alternating layers of fluid and interfaces between these layers. A similar interpretation applies to block copolymers where the composition variable describes the relative local density of the two blocks. However, the same order parameter could equally describe the local number density of the midpoints of a rodlike mesogenic molecule (measured relative to the mean) which can develop a periodic density wave that represents the smectic state.

The governing equations in our model are a convection-diffusion equation
for the scalar order parameter, and the Navier-Stokes equation describing
the hydrodynamics of fluid flow. The stress tensor for the fluid contains not only a Newtonian viscous term but extra contributions arising from local deviations of the order parameter from its state of least free energy. Diffusive relaxation of the order parameter then causes additional dissipation, resulting in an effective viscosity that can be much higher than the intrinsic Newtonian contribution.
Our dynamical equations are solved using a combination of a finite
difference scheme for the convection-diffusion equation, and the
lattice Boltzmann method (LBM)~\cite{Succi} for the Navier-Stokes equation \cite{Marenduzzo07, Henrich10a,  Henrich10b, Henrich11a,Gonnella97}.
We use sliding periodic (Lees-Edwards) boundary conditions for
both the convection-diffusion and Navier-Stokes equations. Note that while these boundary conditions impose a fixed integrated strain rate across the whole system they do not impose a uniform strain rate locally; shear bands can therefore arise if the system so chooses. (As it happens, we do not observe shear bands in the range of shear rates we address here.)

@@

KS: THIS FOLLOWING PARAGRAPH DOESN'T QUITE MAKE SENSE TO ME YET.

Our main finding is that, in two dimensions at least, sheared smectics can show two completely different types of rheological response. At modest rates
of shear, where thermodynamic forces are relatively strong compared with
flow effects,  shear quickly aligns the system into the perfect lamellar phase preferred by thermodynamics,
with only a few defects. The rheology in this regime is quasi-Newtonian, with shear stress linear in the applied shear rate. (There is however a large first normal stress difference.) At higher shear rates, larger thermodynamic forces are generated as layers fail to keep up with the imposed flow. This leads to the nucleation of many new defects, balanced in steady state by annihilation when defects of opposite sign encounter each other. The strain rate $\dot\gamma_c$ needed to enter this nonlinear flow regime depends on thermodynamic and kinetic parameters; a comprehensive exploration of parameter space is not possible, but we give a scaling argument which relates $\dot\gamma_c$ to those parameters through an effective capillary number, formed as the ratio of viscous to thermodynamic stress.
Within the defect proliferation regime, the macroscopic rheology is generically nonlinear, and shows shear thinning. The shear stress appears to vary as a sublinear power law, $\sigma_{xy}\propto\dot{\gamma}^m$, with $m\simeq 0.87$ for typical parameter values, although the dependence of $m$ on the parameters precludes interpretation of this as a universal scaling law.
Such a behaviour is qualitatively reminiscent of the one found experimentally in [REFERENCE], where the numerical value of $m$ was however different. In this strongly non-Newtonian regime we also find that the rheology is history dependent.

@@

The equilibrium phase in a system of volume $V$ is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimises the Landau-Brazovskii free energy functional \cite{Brazovskii75, Gompper}.
%
\begin{eqnarray}
{\cal F}[\phi] &=& \int d^dx f(\phi,\nabla\phi,\nabla^2\phi) = V\bar f\\ &=& \int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}\nonumber\label{FE-functional}
\end{eqnarray}
%
In minimizing this free energy at equilibrium, we ignore thermal fluctuations (treated in \cite{Brazovskii75,Cates89}), thereby adopting a mean-field approximation as in previous literature \cite{Kendon01,Gonnella97, Gonnella98, Xu03, Xu05, Xu06a, Xu06b}. This should be a good approximation far from the ordering transition at which $\phi$ first becomes nonuniform. The uniform state has $\phi = 0$ and hence ${\cal F}= 0$.

The parameter range relevant to lamellar phases has $\kappa < 0$ and $c>0$. Then, if we suppress the quartic term, the free energy density of a sinusoidal lamellar state $\phi({\bf r}) = \phi_k\cos({\bf k}.{\bf r})$ obeys $\bar f = (a+\kappa k^2+ c k^4)\phi_k^2/4$. This is minimized by choosing $k=q\equiv\sqrt{|\kappa|/2c}$; the free energy density is then $\bar f = (a-\kappa^2/4c)\phi_q^2/4$ so that ordering will occur for $a<a_c=\kappa^2/4c$. The amplitude $\phi_q$ is then unbounded without the (positive) quartic term, since $\bar f$ can in that manner become infinitely negative. Restoring the quartic term now causes saturation at $\phi_q\sim \sqrt{4(a_c-a)/3b}$, creating also nonlinear harmonics in the density profile. The final free energy density obeys $\bar f \simeq (a_c-a)^2/b$. By similar arguments, the characteristic elastic modulus associated with microstructural deformations is
%
\begin{equation}
\bar g\sim \frac{(a_c-a)a_c}{4b}
\label{barg}
\end{equation}
%
which is the energy density of the gradient terms only. The terms involving $\phi^2$ and $\phi^4$ in Eq.~\ref{FE-functional} are unaltered by affine deformations which conserve density locally; hence those terms do not contribute to $\bar g$.

The dynamics of the order parameter is governed by an advection-diffusion equation with a diffusive current of $\phi$ dependent on the chemical potential $\mu$, the functional derivative of ${\cal F}$ with respect to $\phi$, as follows:
%
\begin{equation}\label{OPEOM}
\partial_t \phi +{\nabla}({\bm u}\phi) = M \nabla^2 \frac{\delta \cal F}{\delta \phi}=M \nabla^2 \mu 
\end{equation}
%
Here $M$ is a mobility coefficient. A wavevector-dependent linearized collective diffusivity $D_k$ can then be defined via $\dot\phi_k = -k^2D_k\phi_k$, whose value at the ordering wavevector obeys $D_q = M(a-a_c)/2$ (this is negative, signifying the instability). In contrast, an estimate of the molecular diffusivity can be made by considering zero wavevector, setting $b = -a$ (see below) and examining the relaxation equation for small deviations around the uniformly ordered state at $\phi = \pm 1$. This gives $\dot\phi = 2bM\nabla^2\phi$ so that the molecular diffusivity is of order $\tilde D = 2Mb$.

Treating the fluid as incompressible, the evolution of the velocity ${\bm u}$ obeys the Navier-Stokes equation
%
\begin{equation}\label{NSE}
\rho(\partial_t u_\alpha + (u_\beta \nabla_\beta) u_\alpha) = \partial_\beta \sigma_{\alpha \beta}. 
\end{equation}
%
where $\rho$ is the mass density. The stress tensor in this equation reads
%
\begin{equation}
\sigma_{\alpha \beta}=\eta (\partial_\alpha u_\beta + \partial_\beta u_\alpha)-p\delta_{\alpha\beta} - P_{\alpha \beta}. \label{fullstress}
\end{equation}
%
This consists of a Newtonian hydrodynamic part, which is proportional to the background viscosity $\eta$ and the symmetrized velocity gradient, a hydrostatic pressure term, which includes an isotropic stress contribution from the order parameter 
%
\begin{eqnarray}
p&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\nabla}\phi)^2\right)\nonumber\\
& &\quad+c\left(\phi(\nabla^2)^2\phi+\frac{1}{2}(\nabla^2\phi)^2+\partial_{\gamma}\phi\partial_\gamma(\nabla^2\phi)\right)
\end{eqnarray}
%
and finally another term consisting of the order-parameter contribution to the deviatoric stress
\cite{Yang76,Evans79} [REFERENCE OK?]
%
\begin{equation}
P_{\alpha \beta}=\kappa\partial_\alpha\phi\partial_\beta\phi -c \left(\partial_\alpha\phi\partial_\beta(\nabla^2\phi)+\partial_\beta\phi\partial_\alpha(\nabla^2\phi)\right).
\end{equation}

@@

To calibrate the LB length unit we equate $L$ the lamellar wavelength with that of a typical thermotropic smectic: $10$ LBU $= L = 4$nm. Hence the LB length unit is about $0.4$nm; and 1m$= 2.5\times 10^{9}$LBU. Free energy density parameters such as $a_c \simeq 10^{-4}$LBU should then be of order $\alpha^2k_BT/L^3$ where $\alpha =10$ is a molecular aspect ratio. Thus the LB unit of energy density is about $6\times 10^{10}$Pa, i.e. 1 Pa$= 1.6 \times 10^{-11}$ LBU.

@@

As detailed previously, an estimate of the molecular diffusivity is $\tilde D = 2Mb$; we set $M = 0.25$LBU throughout this study, so that $\tilde D$ lies in the range $2.5\times 10^{-5} - 2.5\times 10^{-4}$LBU. Taking a mid-range value of  $\tilde D = 10^{-4}$LBU and using the above conversion factors (1 m$= 2.5\times 10^9$LBU, 1s $= 6\times 10^{12}$LBU) gives $D\simeq 10^{-10}$m$^2$s$^{-1}$ which is a reasonable mesogenic value.

Recall that the mass density of our fluid is unity in LBU. However, combining the above conversion factors for length,  energy density, and time shows the physical and simulation units of mass density to be related via 1 kg m$^{-3} \equiv 1$ Pa s$^2$ m$^{-2} \simeq 10^{-8}$LBU. Thus our simulations address a fluid of mass density $10^8$kg m$^{-3}$ which is $10^5$ times larger than the true value for a typical material. The reason for this is that the LB algorithm uses fluid inertia to update the dynamics and the discretization should therefore involve replacing the real fluid with the densest one possible, while ensuring that the Reynolds number $Re \sim \rho\dot\gamma L^2/\eta$ (which for real smectics is exceedingly small) remains small enough for the simulation to faithfully describe the low $Re$ limit. As discussed elsewhere \cite{Cates04, Cates09} $Re$ values up to about 0.1 are then admissable; our simulations comply with this requirement.

Finally, our shear rates lie in the range 
$10^{-5}-10^{-4}$LBU, so that the (dimensionless) Peclet number, $Pe =\dot\gamma L^2/\tilde D$, which expresses the relative importance of advective to diffusive transport, lies in the range $5-800$. These are moderate to high values where significant structural rearrangement can be expected. However, for thermotropic smectics, given our time unit of 1LBU$=1.6\times 10^{-12}$s, they correspond to strain rates of order $10^6$s$^{-1}$. This lies well above the range typically addressable with rheological experiments, but below that recently given in a theoretical study \cite{Stewart2009}.

The shear rates needed to achieve this range of $Pe$ falls dramatically when lyotropic smectics are considered. For lyotropics one has $L\sim 40$nm; the LB length unit is thus about 4nm. The factor $\alpha$ in the free energy density is now of order unity (since the main contributions are entropic and for flexible membranes there is about one relevant degree of freedom per cube of side $L$) [REFERENCE LYOTROPIC REVIEW ?]. The LB unit of energy density is then about $6\times 10^{5}$Pa, i.e. 1 Pa$= 1.6 \times 10^{-6}$ LBU. Both the viscosity $\eta$ and the molecular diffusivity $\tilde D$ remain comparable (within one order of magnitude) to the thermotropic values; from the viscosity we then determine that 1s$=6\times 10^6$LBU or 1LB timestep $=1.6\times 10^{-7}$s. Accordingly the strain rates relevant to the $Pe$ range addressed here are of order $10^2$ s$^{-1}$ and well within experimental reach, but below that addressed in molecular dynamics simulations \cite{Guo2002,Soddemann2004}. Note that for lyotropics we are effectively simulating a fluid of even higher mass density than for thermotropics; but since the simulated Reynolds numbers are unaffected by the change in parameter mapping, they are still irrelevantly small.
